Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-30 Manuscript/manuscript.tex
Update 'Response of POPC headgroup to bound charge' section

REWRITE:
The headgroup response to bound charge depends on both, the
binding affinity and the response of the headgroup on bound charge.
Thus, the headgroup response to bound charge has to be correctly
described in the model if the concept is used for quantitative studies of
binding affinity. Hence, we first quantify the response of head group order
parameters to the bound charge by using experimental data measured from
mixtures of cationinc surfactants and PC lipids~\cite{scherer89}.

The amount of the bound charge is known excatly in these systems,
because essentially all surfactants with a unit charge locate in the lipid bilayers.
%Thus, the headgroup order parameter response to bound charge can be first
%quantified aginst experimental data before proceeding to binding affinity studies.
%Headgroup order parameter response to the amount of bound charge can
%be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
%All surfactants locate in lipid bilayers due to their amphiphilic nature,
%thus the exact amount of bound charge in the membrane is known in these systems.
%The headgroup order parameter changes as a function of bound charges can be then
%explicitly determined from such systems.

Changes of headgroup order parameters as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESsurf}.
As expected from Eq. \ref{OPchangeEQ}, the head group order parameter response to the bound cation concentration
is approximately linear at least up to $30\%$ mole fraction in experiments and
in both simulation models, original Lipid14 and ECC-lipids.  
\todo{We might want to calculate the slopes from simulations.}

The slope of the head group order parameters response is, however, 
too steep in Lipid14 model indicating that 
its head group response is too sensitive to a bound charge. 
This has to be taken into account when interpreting the
head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.

Also the angle of the headgroup P-N vector with respect to membrane normal
is shown in Fig.~\ref{OrderParameterCHANGESsurf} as a function of bound cations.
As expected~\cite{seelig87}, the headgroup orients more parallel with the membrane
normal with an increasing amount of bound cations. The effect is stronger
in Lipid14 model than in ECC-lipids model, which is in line
with the reduced charge--dipole interactions in the ECC framework  
leading to the lower head group sensitivity to bound charges in ECC-lipids model.

\todo{It might be a good idea to establish a relation between order parameter change
and P-N vector angle by using ECC-lipids model.}

@@

%After achieving accurate head group order parameter response with a cationic surfactant, 
%we advanced to quantifying the response to solutions with varying concentrations of salts, namely \ce{NaCl} and \ce{CaCl_2}.
The headgroup order parameter response to aqueous ionic concentration
can be used to measure ion binding affinity in experiments and simulations,
as discussed in section \ref{section:electrometer} and in previous literature~\cite{seelig87,catte16}.
The headgroup order parameter response of PC lipids to NaCl and CaCl$_2$ concentrations
from experiments (DPPC \cite{akutsu81} and POPC \cite{altenbach84}) and different simulation
models are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
are shown together with the results from Lipid14 simulation
with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    Relative surface excess for the respective simulations are: 
    Lipid14 with \AA{}qvist ions $\Gamma_i^w = XX nm^{-3}$; 
    Lipid14 with Dang ions $\Gamma_i^w = 0.3 nm^{-3}$; 
    ECC-lipids with ECC-ions $\Gamma_i^w = 0.07 nm^{-3}$. 
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{Calculate the surface excess for L14+Aqvist sim.}
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{Resuls with Lipid14+Dang to be added once done -- running with larger box.}
\end{figure}

>>>>

Before proceeding to ion binding affinity studies, we quantify
the response of headgroup order parameters to the amount of 
bound charge by using mixtures of monovalent cationic surfactants
and POPC~\cite{scherer89}. The amount of bound charge per PC 
in these systems is 
given by the molar fraction of cationic surfactants, because essentially all surfactants 
locate in the lipid bilayers. Comparison of headgroup order parameter
changes between simulations and experiments 
as a function of mole fraction of cationic surfactants 
can be then used to validate the sensitivity of
lipid headgroup to bound charge in simulation models.

%depends on both, the
%binding affinity and the response of the headgroup on bound charge.
%Thus, the headgroup response to bound charge has to be correctly
%described in the model if the concept is used for quantitative studies of
%binding affinity. Hence, we first quantify the response of head group order
%parameters to the bound charge by using experimental data measured from
%mixtures of cationinc surfactants and PC lipids~\cite{scherer89}.
%
%Thus, the headgroup order parameter response to bound charge can be first
%quantified aginst experimental data before proceeding to binding affinity studies.
%Headgroup order parameter response to the amount of bound charge can
%be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
%All surfactants locate in lipid bilayers due to their amphiphilic nature,
%thus the exact amount of bound charge in the membrane is known in these systems.
%The headgroup order parameter changes as a function of bound charges can be then
%explicitly determined from such systems.

The headgroup order parameter changes as a function of mole fraction
of cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease of the order parameters is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the headgroup order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

Also the effect of bound cations on the
headgroup P-N vector angle with respect to the membrane normal
is shown in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more parallel to the membrane
normal with an increasing amount of bound cations, as expected~\cite{seelig87}. 
The orientation effect is more pronounced in Lipid14 model than 
in ECC-lipids model, which is in line with the order parameter 
results and reduced charge--dipole interactions in the ECC-lipid model.
Since the response of $\alpha$-order parameter is in very good agreement
with experiments in ECC-lipid model, we can suggest a relation between
order parameter and P-N vector angle changes. The data gives
$\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}$, which can be used
to estimate P-N vector angle changes from experimental results
for $\alpha$ order parameters. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}

@@

The linear decrease of the headgroup order parameters, quantified in previous 
section, can be used to measure the amount of bound aqueous cations in lipid 
bilayers~\cite{seelig87,catte16}. The headgroup order parameter response 
to NaCl and CaCl$_2$ concentrations from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
are shown together with the results from Lipid14 simulation
with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    Relative surface excess for the respective simulations are: 
    Lipid14 with \AA{}qvist ions $\Gamma_i^w = XX nm^{-3}$; 
    Lipid14 with Dang ions $\Gamma_i^w = 0.3 nm^{-3}$; 
    ECC-lipids with ECC-ions $\Gamma_i^w = 0.07 nm^{-3}$. 
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{Calculate the surface excess for L14+Aqvist sim.} \\
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{Resuls with Lipid14+Dang to be added once done -- running with larger box.}
\end{figure}

\subsection{Binding stoichiometry}
